---
title: "Goat Data"
author: "Tanin Rajamand"
date: "2024-12-08"
output:
  pdf_document:
    toc: true
    toc_depth: 5
  html_document:
    toc: true
    toc_depth: 5
---

Our focus will be on analyzing goat-related agricultural data and visualizing it geographically using geospatial tools.

Objective:
To analyze goat-related data, specifically focusing on operations with sales at the county level in the United States. We'll visualize this data on a map to uncover regional trends.

Dataset source: USDA

Questions We’re Answering:
What do goat sales look like across the U.S. at a county level?
How can we clean and prepare this data for geospatial analysis?
How do different regions, like Missouri and Mississippi, compare?
Setting Up the Environment

# Load the Libraries 
before we begin:
```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(viridis)
library(readr)
library(tigris)
library(stringr)

```

sf: Provides tools for working with geospatial data in R. It supports simple features (like polygons, lines, and points), enabling spatial operations like joining data with shapefiles.
ggplot2: A powerful package for creating data visualizations. It allows customization and layering of plots, including geospatial maps.
dplyr: Simplifies data manipulation, like filtering, summarizing, and joining datasets.
viridis: Adds visually appealing, perceptually uniform color scales to your plots. These are great for accessibility (e.g., colorblind-friendly).
readr: Provides fast and friendly tools for reading data files like .csv.
tigris: Downloads and processes Census TIGER/Line shapefiles, which contain geographic and cartographic boundary information for the U.S.
stringr: Simplifies string manipulation tasks like cleaning or formatting text


# Load the Data
lets load the data

```{r}
setwd("~/Lab R/Cattle Supporting Data")
goat <- read_csv("goat.csv")

```
# Looking at the Columns: 
Colnames: 
```{r}
colnames(goat)
```


# Seeing Unique Columns: 
```{r}
# List of columns to check
columns <- c(
  "Program", "Year", "Period", "Week Ending", "Geo Level", "State", 
  "State ANSI", "Ag District", "Ag District Code", "County", "County ANSI", 
  "Zip Code", "Region", "watershed_code", "Watershed", "Commodity", 
  "Data Item", "Domain", "Domain Category", "Value", "CV (%)"
)

# Loop through each column and display unique values
for (col in columns) {
  cat("Unique values in column:", col, "\n")
  print(unique(goat[[col]]))
  cat("\n")
}

```
# Further Information: 
Just an FYI: 
The categories I selected for domain item:
Data Item:

But as you can see in 2022 with census data, we only had 
GOATS - OPERATIONS WITH SALES
GOATS - SALES, MEASURED IN $
GOATS - SALES, MEASURED IN HEAD
GOATS, ANGORA - OPERATIONS WITH SALES
GOATS, ANGORA - SALES, MEASURED IN $
GOATS, ANGORA - SALES, MEASURED IN HEAD
GOATS, MEAT & OTHER - OPERATIONS WITH SALES
GOATS, MEAT & OTHER - SALES, MEASURED IN $
GOATS, MEAT & OTHER - SALES, MEASURED IN HEAD
GOATS, MILK - OPERATIONS WITH SALES
GOATS, MILK - SALES, MEASURED IN $
GOATS, MILK - SALES, MEASURED IN HEAD

But we only chose GOATS - OPERATIONS WITH SALES

ok so lets continue.

lets get rid of the things are either NA or only have one output 

# No NA and No Output: 
Cleaning ensures we’re only working with accurate, relevant data.
```{r}
goat <- subset(goat, select = c(
  "State", "State ANSI", "Ag District", "Ag District Code", "County", "County ANSI", "Data Item","Domain", "Domain Category", "Value", "CV (%)"
))

```

# Load County Shapefiles
We downloaded the data from 2022. 

We'll start by downloading U.S. county boundaries using the tigris package and converting them to an sf object.
This is a special type of data frame with an additional geometry column. Each row represents a county, and the geometry column stores its shape.

This step connects our goat data to geographic boundaries, enabling us to map the trends visually.

counties(): Fetches county-level shapefiles from the U.S. Census Bureau.
cb = TRUE: Simplifies the shapes to make them faster to work with.
resolution = "20m": Sets the level of detail (medium resolution is good for analysis).
st_as_sf(): Converts the shapefiles into an sf object for easy manipulation.
mutate(): Adds or updates columns.
str_to_title(): Converts text to title case (e.g., "new york" → "New York").
sprintf(): Ensures state codes are consistently two characters.
```{r}
# Download county shapefiles for 2021 (simplified geometries for faster plotting)
# we need tirgis here
counties_sf <- counties(cb = TRUE, resolution = "20m", year = 2021) %>%
  st_as_sf()  # Ensure it's an sf object

# Clean county shapefiles
# we need stringr here
counties_sf <- counties_sf %>%
  mutate(
    NAME = str_to_title(NAME),  # Title case for county names
    STATEFP = sprintf("%02d", as.numeric(STATEFP))  # Ensure FIPS is two-character
  )

```

# Filter for the mainland
```{r}
counties_sf_mainland <- counties_sf %>%
  filter(!STATEFP %in% c("02", "15"))
```



# Join the data:
We'll combine the county shapefile with goat sales data. This is where sf shines because it lets us treat geospatial data like regular data frames.
left_join(): Combines the two datasets using matching state and county codes.

```{r}
# Join your data with the filtered counties_sf
joined_data <- counties_sf %>%
  left_join(goat, by = c("STATEFP" = "State ANSI", "COUNTYFP" = "County ANSI"))

# Check the results
nrow(joined_data)  # Should match nrow(counties_sf)
head(joined_data)

```
Notice the added columns from the goat dataset, like Value. The geometry column is preserved, so we can still plot it on a map.


# Remove Hawaii and Alaska: 
lets remove the alaska and hawaii and see if the data matches 
```{r}
# Remove rows with missing data from farmco_mapped
joined_data <- joined_data %>%
  filter(!is.na(Value))  #"Value" is the column containing the data you're interested in

goat <- goat %>%
  filter(!is.na(Value))

# Remove rows for Alaska and Hawaii
goat <- goat %>%
  filter(!State %in% c("HAWAII", "ALASKA"))

joined_data <- joined_data %>%
  filter(!State %in% c("HAWAII", "ALASKA"))

# Check the number of rows after filtering
nrow(goat)
nrow(joined_data)
```
yayy great

# Clean the Data: 
```{r}
# Clean the data using the first approach
joined_data_clean <- joined_data %>%
  mutate(
    Value = as.numeric(gsub("[^0-9.]", "", Value))  # Remove non-numeric characters before conversion
  ) %>%
  filter(!is.na(Value))  # Filter out rows where Value is NA

nrow(goat)
nrow(joined_data)
nrow(joined_data_clean)
sum(is.na(goat$Value))  # Count rows with NA in Value

# Check how many rows in goat are unmatched:
unmatched_goat <- goat %>%
  anti_join(counties_sf, by = c("State ANSI" = "STATEFP", "County ANSI" = "COUNTYFP"))
nrow(unmatched_goat)  # Should match sum(is.na(goat$Value)) if NA rows are unmatched

```
nice

# Map based on Data Item:
lets be reminded of them again:
GOATS - OPERATIONS WITH SALES

geom_sf(): Plots the geospatial data with the Value column dictating the fill color.
scale_fill_viridis_c(): Adds a color scale with a log transformation for better visibility.
labs(): Adds titles and labels.
```{r}
# Filter the cleaned data for the specific Data Item
joined_data_clean_dataitem_opwsales <- joined_data_clean %>%
  filter(`Data Item` == "GOATS - OPERATIONS WITH SALES")

# Map the filtered data
ggplot(data = joined_data_clean_dataitem_opwsales) +
  geom_sf(aes(fill = Value)) +  # Fill counties based on numeric Value
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "gray90",
    trans = "log10",  # Log scale for better color distribution
    direction = -1    # Reverse color scale if needed
  ) +
  theme_minimal() +
  labs(
    title = "County-Level GOATS - OPERATIONS WITH SALES",
    fill = "OPERATIONS WITH SALES"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```
Here’s the final visualization. Each county is colored based on the number of goat operations with sales. Notice the regional patterns—this could reveal opportunities for targeted interventions or support.

# Missouri: 
lets do Missouri: 
```{r}
# Filter the cleaned data for the specific Data Item
joined_data_clean_dataitem_opwsales <- joined_data_clean %>%
  filter(`Data Item` == "GOATS - OPERATIONS WITH SALES", State == "MISSOURI")

# Map the filtered data
ggplot(data = joined_data_clean_dataitem_opwsales) +
  geom_sf(aes(fill = Value)) +  # Fill counties based on numeric Value
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "gray90",
    trans = "log10",  # Log scale for better color distribution
    direction = -1    # Reverse color scale if needed
  ) +
  theme_minimal() +
  labs(
    title = "MISSOURI - County-Level GOATS - OPERATIONS WITH SALES",
    fill = "OPERATIONS WITH SALES"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```

# MISSISSIPPI: 
lets do MISSISSIPPI: 
```{r}
# Filter the cleaned data for the specific Data Item
joined_data_clean_dataitem_opwsales <- joined_data_clean %>%
  filter(`Data Item` == "GOATS - OPERATIONS WITH SALES", State == "MISSISSIPPI")

# Map the filtered data
ggplot(data = joined_data_clean_dataitem_opwsales) +
  geom_sf(aes(fill = Value)) +  # Fill counties based on numeric Value
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "gray90",
    trans = "log10",  # Log scale for better color distribution
    direction = -1    # Reverse color scale if needed
  ) +
  theme_minimal() +
  labs(
    title = "MISSISSIPPI - County-Level GOATS - OPERATIONS WITH SALES",
    fill = "OPERATIONS WITH SALES"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```

________________________

